---
title: "Prevalence of anti-SARS-CoV-2 IgG antibodies in a representative sample from the Canton of Geneva"
output:
  html_document:
    df_print: paged
  pdf_document: default
  word_document: default
params:
  redo: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE)
knitr::opts_knit$set(root.dir = here::here())
```

```{r preamble, include=FALSE}
library(tidyverse)
library(rstan)
library(doMC)
library(gridExtra)

source("scripts/utils.R")

cbbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")

## age cuts
age_cuts <- c(5, 10, 20, 50, 65, 105)

## Stan/JAGS settings, don't need to change
options(mc.cores = 4)
n_chains <- 4
n_iter <- 1500
n_warmup <- 250
p_delta <- 0.99
n_treedepth <- 20

# Output estimate file
out_est_file <- "generated_data/re-seropos.rds"
```

```{r validation data, include=FALSE}
## bring in lab validation data
lab_val_dat <- readxl::read_xlsx("data/20200508 Serology ELISA-IF validation overview_clean.xlsx", sheet = "All Results")

## number of positive controls from validation data
## Meyer et al, 2020
pos_control <- sum(lab_val_dat$Type == "COVID-19 Patient") # 181
## number of negative controls
## Meyer et al, 2020
neg_control <- sum(lab_val_dat$Type == "Neg. Ctrl.") # 176

## true positive rate for cases
control_tp_hi <- lab_val_dat %>%
  filter(Type == "COVID-19 Patient") %>%
  summarize(fp = sum(`ELISA/CE G Results` == "POS")) %>%
  unlist() # 154

control_tp_gva <- lab_val_dat %>%
  filter(Type == "COVID-19 Patient") %>%
  summarize(fp = sum(`ELISA/CE G Result GE` == "POS")) %>%
  unlist() # 143

control_tp_ifa <- lab_val_dat %>%
  filter(Type == "COVID-19 Patient") %>%
  summarize(fp = sum(`ELISA/CE G Result GE` != "NEG" & `rIFA G Result` == "POS")) %>%
  unlist() # 161

## false positive rate for controls (1-specificity)
control_fp_hi <- lab_val_dat %>%
  filter(Type == "Neg. Ctrl.") %>%
  summarize(fp = sum(`ELISA/CE G Results` == "POS")) %>%
  unlist()

control_fp_gva <- lab_val_dat %>%
  filter(Type == "Neg. Ctrl.") %>%
  summarize(fp = sum(`ELISA/CE G Result GE` == "POS")) %>%
  unlist()

control_fp_ifa <- lab_val_dat %>%
  filter(Type == "Neg. Ctrl.") %>%
  summarize(fp = sum(`ELISA/CE G Result GE` != "NEG" & `rIFA G Result` == "POS")) %>%
  unlist()
```

```{r epi data, include = FALSE}
## Geneva incidence data
inc_dat <- read_csv("data/cases_gva.csv") %>%
  mutate(date = as.Date(date, format = "%d/%m/%y"))
inc_dat$incidI <- c(0, diff(inc_dat$ncumul_conf))
inc_dat$incidD <- c(0, diff(inc_dat$ncumul_deceased))

## Geneva ages
gva_ages <- read_csv("data/canton_age_2019.csv") %>%
  mutate_at(
    c("men", "women", "total"),
    function(x) str_replace(x, pattern = " ", repl = "") %>% as.numeric()
  )

## read and fix serology data
sero_dat <- read_csv("generated_data/serocov-pop_data_public.csv") %>%
  rename(household_id = new_household_id) %>% 
  mutate(age_cat = cut(age, age_cuts, right = F)) %>%
  group_by(household_id) %>%
  mutate(
    hh_obs = n(),
    hh_inf = sum(pos) - pos,
    other_inf = hh_inf > 0
  ) %>%
  ungroup() %>%
  droplevels()

wk_ref <- "2"
age_ref <- "[20,50)"
sex_ref <- "0"

## age categories
age_cats <- unique(as.character(sero_dat$age_cat)) %>% sort()

## find population-level proportions of age and sex for sero_dat
gva_age_cats <- gva_ages %>%
  mutate(
    age_med = (age_lower + age_upper) / 2,
    age_cat = cut(age_med, breaks = age_cuts, right = F)
  ) %>%
  filter(age_cat %in% age_cats) %>%
  group_by(age_cat) %>%
  summarize(
    male = sum(men),
    female = sum(women)
  ) %>%
  pivot_longer(
    cols = -age_cat,
    names_to = "Sex",
    values_to = "pop"
  ) %>%
  mutate(
    Sex = ifelse(Sex == "male", 1, 0),
    pct = pop / sum(pop)
  ) %>%
  full_join(
    expand_grid(
      week = unique(sero_dat$week),
      Sex = 0:1
    )
  ) %>%
  arrange(week)

## only the primary member of the household
primary_dat <- sero_dat %>%
  # filter(Codbar==household_id) %>% 
  filter(bus_sante == 1) %>% # is this OK?
  droplevels()

## find population-level proportions of age and sex for primary_dat
primary_age_cats <- gva_ages %>%
  mutate(
    age_med = (age_lower + age_upper) / 2,
    age_cat = cut(age_med, breaks = age_cuts, right = F)
  ) %>%
  filter(age_cat %in% primary_dat$age_cat) %>%
  group_by(age_cat) %>%
  summarize(
    male = sum(men),
    female = sum(women)
  ) %>%
  pivot_longer(
    cols = -age_cat,
    names_to = "Sex",
    values_to = "pop"
  ) %>%
  mutate(
    Sex = ifelse(Sex == "male", 1, 0),
    pct = pop / sum(pop)
  ) %>%
  full_join(
    expand_grid(
      week = unique(primary_dat$week),
      Sex = 0:1
    )
  ) %>%
  arrange(week) %>%
  droplevels()

## set factor levels to reference levels in data
sero_dat$Sex <- relevel(as.factor(sero_dat$Sex), sex_ref)
sero_dat$age_cat <- relevel(sero_dat$age_cat, age_ref)
sero_dat$week <- relevel(as.factor(sero_dat$week), wk_ref)
gva_age_cats$Sex <- relevel(as.factor(gva_age_cats$Sex), sex_ref)
gva_age_cats$age_cat <- relevel(gva_age_cats$age_cat, age_ref)
gva_age_cats$week <- relevel(as.factor(gva_age_cats$week), wk_ref)
primary_dat$Sex <- relevel(as.factor(primary_dat$Sex), sex_ref)
primary_dat$age_cat <- relevel(primary_dat$age_cat, age_ref)
primary_dat$week <- relevel(as.factor(primary_dat$week), wk_ref)
primary_age_cats$Sex <- relevel(as.factor(primary_age_cats$Sex), sex_ref)
primary_age_cats$age_cat <- relevel(primary_age_cats$age_cat, age_ref)
primary_age_cats$week <- relevel(as.factor(primary_age_cats$week), wk_ref)

hh_dat <- sero_dat %>% 
  group_by(household_id) %>% 
  summarize(obs=n(),
            inf=sum(pos),
            head=sum(str_detect(ind_id, "-1")))
```

## Methods

### Statistical model

In this paper, our goal is to estimate the true underlying seroprevalence of the population $\geq 5$ years old as measured each week, $w$ in the of the Canton of Geneva, denoted $p^*_w$ ($w=1,\dots,W=$ `r max(as.numeric(as.character(sero_dat$week)))`).

We start by estimating the probability that each person in the serosurvey is seropositive using a Bayesian logistic regression model that accounts for household clustering, the sensitivity and specificity of the ELISA assay, each individual's age and sex, as well as the week when they were sampled:

\begin{align*}
  x_{i} &\sim Bernoulli(p_i \theta^+ + (1-p_i) * (1-\theta^-)) \\
  logit(p_i) &= \alpha_h + \mathbf{X}_i \boldsymbol{\beta} \\
  \alpha_h &\sim Normal(0, \sigma^2) \\
  x^+ &\sim Binomial(n^+, \theta^+) \\
  x^- &\sim Binomial(n^-, 1-\theta^-)
\end{align*}

where $x_i$ is the result of the IgG ELISA (in primary analyses) for the $i$th person ($i=1,\dots,N=$ `r nrow(sero_dat)`) in the serosurvey.
The sensitivity, $\theta^+$, is determined using $n^+$ RT-PCR positive controls from the lab validation study, of which $x^+$ tested positive.
The specificity, $\theta^-$ , is determined using $n^-$ pre-pandemic negative controls, of which $x^-$ tested positive.
The model estimates of the sensitivity and specificity are shown in Table \@ref(tab:sens-spec).
The probability of observing a diagnostic positive is a function of the true positive rate and the false negative rate with regards to the true underlying probability of seropositivity $p_i$ for that person.
This probability itself is a function of covariates $\mathbf{X}$, which consists of sex, age categories, and week of study, and their coefficients $\boldsymbol{\beta}$, and a random effect for household, $\alpha_h$ ($h=1,\dots, H=$ `r nrow(hh_dat)`), with variance $\sigma^2$.
We used naive priors on all parameters to allow for an exploration of the parameter space.
The priors on the sensitivity and specificity were flat from 0 to 1, equivalent to $Uniform(0,1)$ or $Beta(1,1)$.
We used weak $Normal(0,1)$ priors for the logistic regression coefficients $\boldsymbol{\beta}$.
The prior on the standard deviation of the household effect, $\sigma$, was flat from 0 to infinity (we tested a positive half-Normal and it did not affect estimates).

```{r raw-data}
sero_dat %>%
  mutate(Sex = ifelse(Sex == 0, "Female", "Male")) %>%
  group_by(week, Sex, age_cat) %>%
  summarize(
    obs = n(),
    pos = mean(pos)
  ) %>%
  ggplot(aes(x = as.numeric(week), y = pos, color = age_cat)) +
  geom_point(aes(shape = Sex, size = obs), alpha = 0.5) +
  geom_line(aes(linetype = Sex), alpha = 0.5) +
  scale_x_continuous("Group") +
  scale_y_continuous("Crude pct positive") +
  scale_color_manual("Age", values = cbbPalette[-1]) +
  theme_bw()
```


```{r re-stan, cache=F, eval=T}

if (!file.exists(out_est_file) | params$redo) {
  ## model the overall seropositivity with a random effect for household
  re_seropos <- run_analysis_stan_re(
    model_script = "stan/serosurvey-analysis-re.stan",
    dat = sero_dat,
    analysis = "ei",
    coef_eqn = "Sex + age_cat + week",
    pos_control = pos_control,
    neg_control = neg_control,
    control_tp = control_tp_hi,
    control_fp = control_fp_hi,
    pop_age_cats = gva_age_cats,
    chains = n_chains,
    iter = n_iter,
    warmup = n_warmup,
    control = list(
      adapt_delta = p_delta,
      max_treedepth = n_treedepth
    ),
    save_warmup = F,
    seed = 1,
    redo = params$redo
  )
  
  saveRDS(re_seropos, out_est_file)
  
} else {
  re_seropos <- readRDS(out_est_file)
}

subset_est <- re_seropos$subset_est
```

## Results
```{r tables}
"Number of participants"
nrow(sero_dat)

"Number of households"
nrow(hh_dat)

"Seroprevalence by week"
subset_est %>%
  filter(var == "Week") %>%
  mutate(week = as.factor(as.numeric(val))) %>%
  group_by(sim) %>%
  mutate(coef_val = ifelse(week == 2, NA,
    ifelse(p > p[week == 2], 1, -1)
  )) %>%
  ungroup() %>%
  # mutate(week=relevel(val,"1")) %>%
  left_join(sero_dat %>%
    group_by(week) %>%
    summarize(
      n = n(),
      pos = sum(pos),
      neg = sum(neg),
      ind = sum(ind)
    )) %>%
  group_by(week, n, pos, neg, ind) %>%
  summarize(
    `Seroprevalence (95% CI)` = paste0(
      mean(100 * p) %>%
        formatC(2, format = "f"), " (",
      quantile(100 * p, probs = .025) %>%
        formatC(2, format = "f"), "-",
      quantile(100 * p, probs = .975) %>%
        formatC(2, format = "f"), ")"
    ),
    p = ifelse(is.na(mean(coef_val)), "--",
      min(2 * c(mean(coef_val > 0), mean(coef_val < 0))) %>%
        formatC(4, format = "f")
    )
  ) %>%
  ungroup() %>%
  mutate(
    pos = paste0(pos, " (", formatC(100 * pos / n, 2, format = "f"), "%)"),
    neg = paste0(neg, " (", formatC(100 * neg / n, 2, format = "f"), "%)"),
    ind = paste0(ind, " (", formatC(100 * ind / n, 2, format = "f"), "%)")
  ) %>%
  rename(
    Week = week, Obs = n, `Test positive` = pos, `Test negative` = neg,
    `Indeterminate` = ind
  ) %>%
  knitr::kable(
    align = c("lcccccc"),
    caption = "Seroprevalence by week",
    booktabs = T, linesep = ""
  )

"Relative risks"
subset_est %>%
  filter(var == "Age") %>%
  group_by(sim) %>%
  mutate(rr = ifelse(val == "[20,50)", NA, p / p[val == "[20,50)"])) %>%
  ungroup() %>%
  left_join(sero_dat %>%
    group_by(age_cat) %>%
    summarize(
      n = n(),
      pos = sum(pos),
      neg = sum(neg),
      ind = sum(ind)
    ),
  by = c("val" = "age_cat")
  ) %>%
  bind_rows(subset_est %>%
    filter(var == "Sex") %>%
    group_by(sim) %>%
    mutate(rr = ifelse(val == 0, NA, p / p[val == 0])) %>%
    ungroup() %>%
    mutate(val = ifelse(val == 0, "Female", "Male")) %>%
    left_join(sero_dat %>%
      mutate(val = ifelse(Sex == 1, "Male", "Female")) %>%
      group_by(val) %>%
      summarize(
        n = n(),
        pos = sum(pos),
        neg = sum(neg),
        ind = sum(ind)
      ))) %>%
  group_by(var, val, n, pos, neg, ind) %>%
  summarize(
    `Relative risk (95% CI)` = ifelse(is.na(mean(rr)), "--",
      paste0(
        mean(rr, na.rm = T) %>%
          formatC(3, format = "f"),
        " (", quantile(rr, probs = .025, na.rm = T) %>%
          formatC(3, format = "f"), "-",
        quantile(rr, probs = .975, na.rm = T) %>%
          formatC(3, format = "f"), ")"
      )
    ),
    p = ifelse(is.na(mean(rr)), "--",
      min(2 * c(
        mean(rr > 1, na.rm = T),
        mean(rr < 1, na.rm = T)
      )) %>%
        formatC(4, format = "f")
    )
  ) %>%
  ungroup() %>%
  mutate(
    pos = paste0(pos, " (", formatC(100 * pos / n, 2, format = "f"), "%)"),
    neg = paste0(neg, " (", formatC(100 * neg / n, 2, format = "f"), "%)"),
    ind = paste0(ind, " (", formatC(100 * ind / n, 2, format = "f"), "%)")
  ) %>%
  rename(
    `Test positive` = pos, `Test negative` = neg, `Indeterminate` = ind,
    Obs = n, Category = val
  ) %>%
  select(-var) %>%
  knitr::kable(
    align = c("lcccccc"), booktabs = T,
    linesep = "",
    caption = "Relative risks of seropositivity by age and sex"
  )
```

